RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-08-22 background.tex
Major update section 1, some changes section 2.4, references added

For their experiments, Schriek et al. \cite{SchriekEtal2016} gave the traffic simulator assignment to 12 groups of students in a Software Architecture course at MSc level. They asked the groups to record their two-hour design session, and the recordings were subsequently transcribed (see Appendix~\ref{sect:transcripts:excerpts}). We used these transcripts to develop our argument schemes and critical questions (Section \ref{sect:gmas}), and we also use the traffic simulator case for simple (Figures \ref{fig:example-small}, \ref{fig:args}) and more extensive examples (Section \ref{sect:gmas}).

@@

Figure~\ref{fig:example-small} illustrates a simplified GRL diagram from the traffic simulator design exercise (a full example is shown in Figure~\ref{fig:transcripts:grl} and will be discussed in Section~\ref{sect:gmas}). %MvZ: Since the example no longer matches the full figure I reworded it slightly in the sentence above.
An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system or the system itself (\texttt{Traffic Simulator}, Figure~\ref{fig:example-small}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \texttt{Improve Simulation}) are often related to non-functional requirements, whereas goals (such as  \texttt{Simulate Traffic}) are related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:example-small}, some of the tasks are  \texttt{Create new cars} and \texttt{Keep same cars} which are alternative way of achieving the goal \texttt{Generate Cars}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g., \texttt{Car Objects}). \todo{Floris}{Sepideh}{Belief elements are not explained and not in the example. I think it would be good to include them, as beliefs allow one to give some sort of rationale for e.g. goals (in fact, one of the reviewers of one of our previous papers argued that beliefs can also be used to capture reasons).}

Different links connect the elements in a GRL model. AND, IOR (Inclusive OR), and XOR (eXclusive OR) decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:example-small}, the goal \texttt{Generate cars} is XOR-decomposed to the tasks \texttt{Create new cars} and \texttt{Keep same cars}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Task  \texttt{Create new cars} has a \emph{help} qualitative contribution to the task \texttt{Dynamic Simulation}, and a \emph{hurt} qualitative contribution to the task \texttt{Static Simulation}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors. Here, the goal \texttt{Generate Cars} depends on the resource \texttt{Car Objects}. %SG: I added it to the model but we can also remove it altogether. 
%\todo{Marc}{Sepideh}{Shall we add a dependency relation as well? They don't play a big role in our work..}
%MvZ: It is a bit strange because we say dependencies are between actors but our examples does show this. Perhaps better to leave it out?

@@

Reasoning about which goals to pursue and actions to take is often referred to as \emph{practical reasoning}, and has been studied extensively in philosophy (e.g. \cite{Raz1978-RAZPR,walton1990}) and artificial intelligence \cite{bratman1987,atkinson2007}. One approach is to capture practical reasoning with argumentation or argument schemes~\cite{walton1990}. Applying an argument scheme results in an argument in favor of, for example, taking an action. This argument can then be tested with critical questions about, for instance, whether the action is possible given the situation, and a negative answer to such a question leads to a counterargument to the original argument for the action.

@@

Here, $G$, $A$, and $S$ are variables, which can be instantiated with concrete goals, actions, and softgoals to provide a specific practical argument. For example, a concrete argument about the traffic simulator (cf. Figure \ref{fig:example-small}) is as follows: 
\begin{itemize}
\item[] We have goal \texttt{Simulate Traffic},
\item[] Performing action \texttt{Dynamic Simulation} realizes goal \texttt{Simulate Traffic}, 
\item[] Which contributes positively to the softgoal \texttt{Improve Simulation} 
\item[] \textit{Therefore} 
\item[] We should perform action \texttt{Dynamic Simulation}.
\end{itemize}

Practical reasoning is defeasible, in that conclusions which are at one point acceptable can later be rejected because of new information. Atkinson \emph{et al.}~\cite{atkinson2007} define a set of critical questions that point to typical ways in which a practical argument can be criticized by, for example, questioning the validity of the elements in the scheme or the connections between the elements. Some examples of critical questions are as follows.

@@

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 1: if we find that the task \texttt{Dynamic Simulation} actually does not realize goal \texttt{Simulate Traffic}, we can form a counter-argument. Another way to criticize an argument for an action is to suggest an alternative action that realizes the same goal (question 2). For example, we can argue that task \texttt{Static Simulation} also realizes the goal \texttt{Simulate Traffic}. It is also possible that performing an action has a negative side effect (critical question 3). For example, while the action \texttt{Create New Cars} realizes the goal \texttt{Generate Cars}, it has a negative side effect, namely hurting \texttt{Static Simulation} (cf. Figure \ref{fig:example-small}).

In argumentation, counterarguments are said to \emph{attack} the original arguments (and sometimes vice versa). In the work of Atkinson et al.~\cite{atkinson2007}, arguments and their attacks are captured as an \emph{argumentation framework} of arguments and attack relations. Given an argumentation framework, once can compute which arguments are accepted and which are rejected using different argumentation semantics~\cite{Dung1995}\footnote{Full definitions of argumentation frameworks and semantics will be given in section \ref{sect:gmas}. In this section, we will briefly discuss the intuitions behind these semantics.}. Figure \ref{fig:pras:example} shows an argumentation framework with three arguments from the traffic simulation example, where arguments are rendered as boxes and attack relations as arrows. There is an argument (A1) for \texttt{Keep Same Cars}, an argument (A2) for \texttt{Create New Cars}, and an argument A3 that \texttt{Create New Cars} has a negative effect on \texttt{Static Simulation} (A3). The two alternative PRAS instantiations are A1 and A2. These arguments mutually attack each other, denoted by the double-headed arrow between A1 and A2. Argument A3 attacks A2, as it points to a negative side-effect of \texttt{Create new cars}.

\begin{figure}[ht!]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (att3) [argNodeIN] at (-0.5,-3.5) {\argtext{A1}{Action \texttt{Keep Same Cars} realizes goal \texttt{Generate Cars}\\ \emph{Therefore} we should perform \texttt{Keep Same Cars}}};
        \node[minimum size=1cm] (att1) [argNodeOUT] at (-0.5,0) {\argtext{A2}{Action \texttt{Create New Cars} realizes goal \texttt{Generate Cars}\\ \emph{Therefore} we should perform \texttt{Create New Cars}}};
        \node[minimum size=1cm] (att2) [argNodeIN] at (3.5,0) {\argtext{A3}{\texttt{Create Same Cars} has a negative effect on \texttt{Static Simulation}}};
         \path
    (att2) edge [attackLink] (att1)
    (att1) edge [attackLink,<->] (att3);
\end{tikzpicture}
\caption{Arguments and attacks in the traffic simulation example.}
\label{fig:pras:example}
\end{figure}

Given an argumentation framework, the acceptability of arguments can be determined according to the appropriate argumentation semantics. The intuition is that an argument is acceptable if it is \emph{undefeated}, that is, any argument that attacks it, is itself defeated. In the argumentation framework in Figure~\ref{fig:pras:example}, argument A3 is undefeated because it has no attackers. This makes A2 defeated, because one of its attackers, A3, is undefeated. A1 is then also undefeated, since its only attacker, A2, is defeated by A3. Thus, the set of undefeated (acceptable) arguments given the argumentation framework in Figure~\ref{fig:pras:example} is $\{$A1, A3$\}$.

@@

The usefulness of argument schemes, and PRAS in particular, for the analysis of practical reasoning situations has been shown in different areas such as e-democracy~\cite{cartwright2009IS}, law~\cite{atkinson2005legal}, planning \cite{medellin2013planning} and choosing between safety critical actions \cite{tolchinsky2012deliberation}. In this article, we aim at capturing the stakeholder's discussions as formal argumentation based on PRAS to decide whether intentional elements and their relationships are shown in the resulting goal model. This approach provides a rationalization to the elements of the goal model in terms of underlying arguments, and furthermore, it helps understanding why certain other elements have been accepted and others have been rejected.

Argumentation schemes and their associated critical questions are very well suited for modeling discussions about a goal model: as Murukannaiah et al.~\cite{murukannaiah2015} have shown, they can guide users in systematically deriving conclusions and making assumptions explicit. This can also be seen from the obvious similarities between PRAS (actions, goals, values) and GRL (tasks, goals, softgoals) in the example above.

However, there are also some differences between PRAS and GRL. Not all elements and relationships of GRL fit into PRAS. For instance, PRAS does not have a notion of ``resource'', and many of the relationships of GRL do not occur in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson can actually apply to GRL. Therefore, we develop our own set of argument schemes and critical questions in Section~\ref{sect:gmas} by analyzing transcripts of discussions about the traffic simulator. Before doing so, we give an overview of our framework in the next section.

>>>>

For their experiments, Schriek et al. \cite{SchriekEtal2016} gave the traffic simulator assignment to 12 groups of students in a Software Architecture course at MSc level. They asked the groups to record their design session, and the recordings were subsequently transcribed (see Appendix~\ref{sect:transcripts:excerpts}). We used three of these transcripts as an extensive case study along which we developed our argument schemes and critical questions (Sections \ref{sect:gmas} and \ref{sect:examples}). Furthermore, we also use the traffic simulator case for a simple  running example in this section (Figures \ref{fig:example-small}, \ref{fig:pras:example}).

@@

Figure~\ref{fig:example-small} illustrates a simplified GRL diagram from the traffic simulator design exercise. An actor (\includegraphics[scale=1]{img/actor}) represents a stakeholder of a system or the system itself (\texttt{Traffic Simulator}, Figure~\ref{fig:example-small}). Actors are holders of intentions; they are the active entities in the system or its environment who want goals to be achieved, tasks to be performed, resources to be available, and softgoals to be satisfied. Softgoals (\includegraphics[scale=1]{img/softgoal}) differentiate themselves from goals (\includegraphics[scale=1]{img/goal}) in that there is no clear, objective measure of satisfaction for a softgoal whereas a goal is quantifiable, often in a binary way. Softgoals (e.g. \texttt{Improve Simulation}) are often related to non-functional requirements, whereas goals (such as  \texttt{Simulate Traffic}) are related to functional requirements. Tasks (\includegraphics[scale=1]{img/task}) represent solutions to (or operationalizations of) goals and softgoals. In Figure~\ref{fig:example-small}, we have the two tasks \texttt{Static Simulation} and \texttt{Dynamic Simulation}: if the system can perform both a static and a dynamic simulation, it can achieve goal \texttt{Simulate Traffic}. In order to be achieved or completed, softgoals, goals, and tasks may require resources (\includegraphics[scale=1]{img/resource}) to be available (e.g., \texttt{Car Objects}). \todo{Floris}{Sepideh}{Belief elements are not explained and not in the example. I think it would be good to include them, as beliefs allow one to give some sort of rationale for e.g. goals (in fact, one of the reviewers of one of our previous papers argued that beliefs can also be used to capture reasons).}

Different links connect the elements in a GRL model. AND, IOR (Inclusive OR), and XOR (eXclusive OR) decomposition links (\includegraphics[scale=1]{img/decomposition}) allow an element to be decomposed into sub-elements. In Figure~\ref{fig:example-small}, the goal \texttt{Generate cars} is XOR-decomposed to the tasks \texttt{Create new cars} and \texttt{Keep same cars}, as they are alternative ways of achieving the goal \texttt{Generate Cars}. Contribution links (\includegraphics[scale=1]{img/contribution}) indicate desired impacts of one element on another element. A contribution link has a qualitative contribution type or a quantitative contribution. Task  \texttt{Create new cars} has a \emph{help} qualitative contribution to the task \texttt{Dynamic Simulation}, and a \emph{hurt} qualitative contribution to the task \texttt{Static Simulation}. Dependency links (\includegraphics[scale=1]{img/dependency}) model relationships between actors or resources. Here, the goal \texttt{Generate Cars} depends on the resource \texttt{Car Objects}. %SG: I added it to the model but we can also remove it altogether. 
%\todo{Marc}{Sepideh}{Shall we add a dependency relation as well? They don't play a big role in our work..}
%MvZ: It is a bit strange because we say dependencies are between actors but our examples does show this. Perhaps better to leave it out?

@@

Reasoning about which goals to pursue and actions to take is often referred to as \emph{practical reasoning}, and has been studied extensively in philosophy (e.g. \cite{Raz1978-RAZPR,walton1990}) and artificial intelligence \cite{bratman1987,atkinson2007}. One approach is to capture practical reasoning with argument schemes~\cite{walton1990}. Applying an argument scheme results in an argument in favor of, for example, taking an action. This argument can then be tested with critical questions about, for instance, whether the action is possible given the situation, and a negative answer to such a question leads to a counterargument to the original argument for the action.

@@

Here, $G$, $A$, and $S$ are variables, which can be instantiated with concrete goals, actions, and softgoals to provide a specific practical argument. For example, a concrete argument about the traffic simulator is as follows: 
\begin{itemize}
\item[] We have goal \texttt{Simulate Traffic},
\item[] Performing action \texttt{Static Simulation} realizes goal \texttt{Simulate Traffic}, 
\item[] Which contributes positively to the softgoal \texttt{Improve Simulation} 
\item[] \textit{Therefore} 
\item[] We should perform action \texttt{Static Simulation}.
\end{itemize}

Practical reasoning is defeasible, in that conclusions which are at one point acceptable can later be rejected because of new information. Atkinson et al.~\cite{atkinson2007} define a set of critical questions that point to typical ways in which a practical argument can be criticized by, for example, questioning the validity of the elements in the scheme or the connections between the elements. Some examples of critical questions are as follows.

@@

These critical questions can point to new arguments that might counter the original argument. Take, for example, critical question 1: if we find that performing a \texttt{Static Simulation} actually does not realize goal \texttt{Simulate Traffic}, we can form a counter-argument. Another way to criticize an argument for an action is to suggest an alternative action that realizes the same goal (question 2). For example, we can argue that performing a \texttt{Dynamic Simulation} also realizes the goal \texttt{Simulate Traffic} on its own. Finally, it is possible that performing an action has a negative side effect (critical question 3). For example, while the action \texttt{Create New Cars} realizes the goal \texttt{Generate Cars}, it has a negative side effect, namely hurting \texttt{Static Simulation}: having the simulation constantly create new cars is a functionality that does not allow for a static simulation.

In argumentation, counterarguments are said to \emph{attack} the original arguments (and sometimes vice versa). In the work of Atkinson et al.~\cite{atkinson2007}, arguments and their attacks are captured as an \emph{argumentation framework} of arguments and attack relations. Given an argumentation framework, we can compute which arguments are accepted and which are rejected using different argumentation semantics~\cite{Dung1995}\footnote{Formal definitions of argumentation frameworks and semantics will be given in section \ref{sect:gmas}. In this section, we will briefly discuss the intuitions behind these concepts.}. Figure \ref{fig:pras:example} shows an argumentation framework with three arguments from the traffic simulation example, where arguments are rendered as boxes and attack relations as arrows. There are two slightly simplified practical reasoning arguments based on PRAS: argument A1 for \texttt{Keep Same Cars} and argument A2 for \texttt{Create New Cars}. Argument A2 proposes an alternative way of realizing the same goal \texttt{Generate Cars} with respect to argument A1 and vice versa (cf. critical question 2), so A1 and A2 mutually attack each other, denoted by the double-headed arrow between A1 and A2. Argument A3 says that \texttt{Create New Cars} has a negative effect on \texttt{Static Simulation}, so A3 attacks A2, as it points to a negative side-effect of \texttt{Create new cars} (critical question 3).

\begin{figure}[ht!]
\centering
\begin{tikzpicture}
        \node[minimum size=1cm] (att3) [argNodeIN] at (-0.5,-3.4) {\argtext{A1}{Action \texttt{Keep Same Cars} realizes goal \texttt{Generate Cars}\\ \emph{\underline{Therefore}} we should perform \texttt{Keep Same Cars}}};
        \node[minimum size=1cm] (att1) [argNodeOUT] at (-0.5,0) {\argtext{A2}{Action \texttt{Create New Cars} realizes goal \texttt{Generate Cars}\\ \emph{\underline{Therefore}} we should perform \texttt{Create New Cars}}};
        \node[minimum size=1cm] (att2) [argNodeIN] at (3.5,0) {\argtext{A3}{\texttt{Create New Cars} has a negative effect on \texttt{Static Simulation}}};
         \path
    (att2) edge [attackLink] (att1)
    (att1) edge [attackLink,<->] (att3);
\end{tikzpicture}
\caption{Arguments and attacks in the traffic simulation example.}
\label{fig:pras:example}
\end{figure}

Given an argumentation framework, the acceptability of arguments can be determined according to the appropriate argumentation semantics~\cite{Dung1995}. The intuition is that an argument is acceptable if it is \emph{undefeated}, that is, any argument that attacks it, is itself defeated. In the argumentation framework in Figure~\ref{fig:pras:example}, argument A3 is undefeated because it has no attackers. This makes A2 defeated (indicated by the lighter grey color of A2), because its attacker A3 is undefeated. A1 is then also undefeated, since its only attacker, A2, is defeated by A3. Thus, the set of undefeated (acceptable) arguments given the argumentation framework in Figure~\ref{fig:pras:example} is $\{$A1, A3$\}$.

@@

In previous work on the RationalGRL framework~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we have explored one way of combining PRAS and GRL. In this work, which takes a similar approach to \cite{Jureta:RE2008}, argument diagrams are translated to GRL models (an automatic translation tool is discussed in~\cite{vanZee-etal:comma2016}). The argument diagrams are complex practical reasoning arguments, where the premises are the goals and the conclusion is some action we need to perform to realize those goals. Thus, we have essentially two complex diagrams, a practical reasoning argument diagram and a GRL goal diagram, and a mapping between them.

Connecting argumentation and goal modelling by providing a mapping between two types of diagram was, in our opinion, ultimately an unsatisfying solution given the problems and requirements described in Section \ref{sect:introduction}. One problem is that the argument diagram is at least as complex as the GRL diagram, so any stakeholder trying to understand the discussion thus far has to parse two complex diagrams containing goals, alternative solutions, tasks, and so forth. Furthermore, in our previous work we did not provide the specific critical questions for goal models (see Section \ref{sect:gmas}). This meant that any counterargument had to be constructed from scratch, as no guidance was given as to possible ways to criticize a GRL model. So the previous iterations of the RationalGRL framework violated requirement 1: argument diagrams do not closely mirror the actual discussions of stakeholders in which ideas are proposed and challenged. Furthermore, not having specific argument schemes and critical questions for goal modelling makes it hard to develop a guiding methodology for the use of argumentation in goal modelling.

In the current version of GRL, we do not literally take the Practical Reasoning Argument Scheme and its critical questions - this would be impossible given that there are elements of the GRL language, such as resources, which cannot be found in PRAS. Furthermore, it is not directly clear whether the critical questions as proposed by Atkinson and Bench-Capon~\cite{atkinson2007} actually apply to GRL models. Therefore, we develop our own set of argument schemes and critical questions in Section~\ref{sect:gmas} by analyzing transcripts of discussions about the traffic simulator. Before doing so, we give an overview of our framework in the next section.
